---
title: "quadrupen: Sparsity by Worst-Case Quadratic Penalties"
format: gfm
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/jchiquet/quadrupen/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jchiquet/quadrupen/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = "man/figures/")

```

## Description

Fits classical sparse regression models with efficient active set
algorithms by solving quadratic problems. Also provides a few methods
for model selection purpose (cross-validation, stability selection).

## Installation

```{r install, eval=FALSE}
devtools::install_github("jchiquet/quadrupen").
```

Many examples and demos can be found in the `inst` repositories.  Here is a more illustrative one: 

## Example: structured penalized regression

This example is extracted form chapter 1 of my [habilitation](https://tel.archives-ouvertes.fr/tel-01288976/). You can get more insight by reading pages 65-66, 70-71. 

Ok first load the package

```{r}
library(quadrupen)
```

### A toy data set advocating for structured regularization

See pages 29--30 in my [habilitation](https://tel.archives-ouvertes.fr/tel-01288976/). Code for additional function is given by the end of the present document (Appendix).

```{r, echo=FALSE}
chol.uniform <- function(p,rho) {

  q <- 0     ## sum_k^(i-1) c.k^2
  cii <- 1   ## current value of the diagonal
  c.i <- rho ## current value of the off-diagonal term
  Cii <- cii ## vector of diagonal terms
  C.i <- c.i ## vector of off-diagonal terms
  for (i in 2:p) {
    q <- q + c.i^2
    cii <- sqrt(1-q)
    c.i <- (rho-q)/cii
    Cii <- c(Cii,cii)
    C.i <- c(C.i,c.i)
  }
  return(list(cii=Cii, c.i=C.i[-p]))
}

rlm <- function(x,beta,mu=0,r2=NULL,sigma=1) {
  n <- nrow(x)
  if (!is.null(r2))
      sigma  <- as.numeric(sqrt((1-r2)/r2 * t(beta) %*% cov(x) %*% beta))
  epsilon <- rnorm(n) * sigma
  y <- mu + x %*% beta + epsilon
  r2 <- 1 - sum(epsilon^2) / sum((y-mean(y))^2)
  return(list(y=y,sigma=sigma))
}

## Blockwise structure
## the length of rho determine the number of blocs
## each group must be > 1 individual
rPred.block <- function(n, p, sizes=rmultinom(1,p,rep(p/K,K)), rho=rep(0.75,4)) {

  K <- length(rho)
  stopifnot(sum(sizes) == p | length(rho) != length(sizes))
  Cs <- lapply(1:K, function(k) chol.uniform(sizes[k],rho[k]))
  
  rmv <- function() {
      return(
      unlist(lapply(1:K, function(k) {
        z <- rnorm(sizes[k])
        Cs[[k]]$cii*z + c(0,cumsum(z[-sizes[k]]*Cs[[k]]$c.i))
      }))
      )
  }
  return(t(replicate(n, rmv(), simplify=TRUE)))
}
```

We draw data from linear regression where the regression parameters are defined group-wise. The corresponding regression correlated according to the same pattern.

```{r predictor}
n <- 200
p <- 192

## model settings: block wise
mu <- 0
group  <- c(p/4,p/8,p/4,p/8,p/4)
labels <- factor(rep(paste("group", 1:5), group))
beta   <- rep(c(0.25,1,-0.25,-1,0.25), group)
x <- rPred.block(n, p, sizes = group, rho=c(0.25, 0.75, 0.25, 0.75, 0.25))
```

Indeed, the correlation structure between the regressors exhibits a strong pattern:
```{r echo=FALSE}
Matrix::image(Matrix::Matrix(cor(x)))
```

We draw the response variable by fixing the variance of the noise ratio to get an $R^2$ equal to $0.8$.
```{r}
dat <- rlm(x,beta,mu,r2=0.8)
y <- dat$y
sigma <- dat$sigma
```

### Regularization without prior knowledge

No we try the available penalized regression method in `quadrupen` to fit the regularization path to this data set

#### Ridge regression  ($\ell_2$ penalty)

```{r ridge_nostruct}
plot(ridge(x,y, intercept=FALSE), labels=labels)
```

#### Lasso  ($\ell_1$ penalty)

```{r lasso_nostruct}
plot(lasso(x,y, intercept=FALSE), labels=labels)
```

#### Elastic-net  ($\ell_1+\ell_2$ penalty)

```{r enet_nostruct}
plot(elastic.net(x,y, lambda2=1, intercept=FALSE), labels=labels)
```

#### Bounded regression ($\ell_\infty$ penalty)

```{r breg_nostruct}
plot(bounded.reg(x,y, lambda2=0, intercept=FALSE), labels=labels)
```

#### Ridge + Bounded regression ($\ell_\infty+\ell_2$ penalty)

```{r bregridge_nostruct}
plot(bounded.reg(x,y, lambda2=5, intercept=FALSE), labels=labels)
```

#### Lava (sparse + dense signal decomposition)

```{r lava_nostruct}
out_lava <- lava(x,y, lambda2=1, intercept=FALSE)
out_lava$plot_path(component = "sparse", labels=labels)
out_lava$plot_path(component = "dense", labels=labels)
```

### Regularization with smooth prior knowledge

Now let use define the graph associated with the groups of the regressor and compute the Graph Laplacian. We add a small value on the diagonal ton ensure strict positive definiteness (in he future, this matrix should preferentially be defined with a graph).

```{r prior}
A <- Matrix::bdiag(lapply(group, function(s) matrix(1,s,s))) ; diag(A) <- 0
L <- -A; diag(L) <- Matrix::colSums(A) + 1e-2
```

Now, we run all the method having a ridge-like regularization by replacing the ridge penalty $\| \code \|_2^2$ with $\| \code \|_{\mathbf{L}}^2$ to enforce some structure in the regularization: the solution paths look a lot more convincing.

#### Structured/Generalized Ridge regression  ($\ell_2$ penalty)

```{r ridge_struct}
plot(ridge(x,y, lambda = 10^seq(4,-1,len=100), struct = L, intercept=FALSE), labels=labels)
```


#### Structured/Generalized Elastic-net  ($\ell_1+\ell_2$ penalty)

```{r enet_struct}
plot(elastic.net(x,y, struct = L, lambda2=1, intercept=FALSE), labels=labels)
```

#### Bounded regression  + structured/Generalized Ridge ($\ell_\infty+\ell_2$ penalty)

```{r bregridge_struct}
plot(bounded.reg(x,y, struct = L, lambda2=5, intercept=FALSE), labels=labels)
```

#### Lava (sparse + structured dense)

```{r lava_struct}
out_lava <- lava(x,y, lambda2=1, struct = L, intercept=FALSE)
out_lava$plot_path(component = "sparse", labels=labels)
out_lava$plot_path(component = "dense", labels=labels)
```

### Regularization with hard prior knowledge

We can use the correlation structure as a proxy for forming groups of variables, and use group-sparse regularization. We offer three variants of group sparse regularization in `quadrupen`: standard group-lasso ($\ell_1/\ell_2$ mixed norm), type 2 group Lasso ( ($\ell_1/\ell_\infty$ mixed norm)) and the [cooperative-lasso](doi.org/10.1214/11-AOAS520), an original proposition.

```{r grp-data}
group <- rep(1:5, c(25,10,25,10,25)) 
```

#### Group-Lasso (group sparse L1/L2)

```{r grplasso-l1l2}
plot(group.lasso(x, y, group, type = "l2", lambda2=0, intercept=FALSE), labels=labels)
```

#### Group-Lasso (group sparse L1/LInf)

```{r grplasso-l1linf}
plot(group.lasso(x, y, group, type = "linf", lambda2=0, intercept=FALSE), labels=labels)
```

#### Cooperative-Lasso (sign-coherent group lasso)

```{r cooplasso}
plot(group.lasso(x, y, group, type = "coop", lambda2=0, intercept=FALSE), labels=labels)
```

### Regularization mixing smooth and hard prior knowledge

We let the possibility to add structured-$\ell_2$ penalty to this mixed-group penalties, in order to introduce additional smoothing prior in the regularization, just like in the structured version of the Elastic-Net and Ridge regression

#### Group-Lasso L2 + Structured ElasticNet (group sparse L1/L2 + structured L2)

```{r grpenet-l1l2}
plot(group.lasso(x, y, group, type = "l2", lambda2 = 2, struct = L, intercept=FALSE), labels=labels)
```

#### Group-Lasso LInf + Structured ElasticNet (group sparse L1/LInf + structured L2)

```{r grpenet-l1linf}
plot(group.lasso(x, y, group, type = "linf", lambda2 = 2, struct = L, intercept=FALSE), labels=labels)
```

#### Cooperative-Lasso + Structured ElasticNet (sign-coherent group lasso + structured L2)

```{r coopenet}
plot(group.lasso(x, y, group, type = "coop", lambda2 = 2, struct = L, intercept=FALSE), labels=labels)
```

## Appendix: functions for data generation

```{r appendix_func_data, eval=FALSE}
chol.uniform <- function(p,rho) {

  q <- 0     ## sum_k^(i-1) c.k^2
  cii <- 1   ## current value of the diagonal
  c.i <- rho ## current value of the off-diagonal term
  Cii <- cii ## vector of diagonal terms
  C.i <- c.i ## vector of off-diagonal terms
  for (i in 2:p) {
    q <- q + c.i^2
    cii <- sqrt(1-q)
    c.i <- (rho-q)/cii
    Cii <- c(Cii,cii)
    C.i <- c(C.i,c.i)
  }
  return(list(cii=Cii, c.i=C.i[-p]))
}

rlm <- function(x,beta,mu=0,r2=NULL,sigma=1) {
  n <- nrow(x)
  if (!is.null(r2))
      sigma  <- as.numeric(sqrt((1-r2)/r2 * t(beta) %*% cov(x) %*% beta))
  epsilon <- rnorm(n) * sigma
  y <- mu + x %*% beta + epsilon
  r2 <- 1 - sum(epsilon^2) / sum((y-mean(y))^2)
  return(list(y=y,sigma=sigma))
}

## Blockwise structure
## the length of rho determine the number of blocs
## each group must be > 1 individual
rPred.block <- function(n, p, sizes=rmultinom(1,p,rep(p/K,K)), rho=rep(0.75,4)) {

  K <- length(rho)
  stopifnot(sum(sizes) == p | length(rho) != length(sizes))
  Cs <- lapply(1:K, function(k) chol.uniform(sizes[k],rho[k]))
  
  rmv <- function() {
      return(
      unlist(lapply(1:K, function(k) {
        z <- rnorm(sizes[k])
        Cs[[k]]$cii*z + c(0,cumsum(z[-sizes[k]]*Cs[[k]]$c.i))
      }))
      )
  }
  return(t(replicate(n, rmv(), simplify=TRUE)))
}
```

